Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2

### 2023-08-21 go3.Rmd
third attempt

---
title: "Estimating differences in scale for high throughput sequencing analysis with ALDEx2"
shorttitle: "Scale ALDEx2"
author:
- name: Greg Gloor, Michelle Pistner Nixon, Justin Silverman
  affiliation: Dep't of Biochemistry, University of Western Ontario, Penn State
  email: ggloor@uwo.ca
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  BiocStyle::pdf_document: default
package: ALDEx2
abstract: |
    Introduction to scale simulation and FDR correction with ALDEx2.
---
# scale simulation for the no-math crowd

Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis

@@

High throughput sequencing (HTS) analysis is commonly used to read out many different types of experimental designs; bulk and single-cell transcriptomics, amplicon and shotgun metagenomics, in vitro selection experiments, and more. There are many analysis tools that are purpose-built for each experimental design and that work well when all assumptions are fulfilled. However, different tools make different assumptions about the underlying data and when the tools fail, they usually fail without error.

Despite the variety, all tools and analyses suffer from two common issues. First, that many features are statistically significantly different between groups even though the measured difference is very small. Second, that the results can be skewed if the underlying dataset has an asymmetric distribution between the groups. Here we show that both of these issues can be resolved by modelling the size, or scale, of the data in the underlying environment.

It is instructive to understand how the data are generated. HTS starts by making a 'library' which is a fixed-size sample from the environment. The library is usually combined with other libraries through 'multiplexing' where  the goal is to combine the libraries such that each library contributes equivalent numbers of molecules to the pool. Finally, a fixed number of molecules is sampled from the pooled samples and loaded onto the instrument. Thus, the process of sequencing is akin to taking a poll (making a library), combining polls (multiplexing into a pool) and taking a poll of the pool. It should be clear that this process removes any relationship between actual numbers in the environment and the actual number of molecules actually sequenced. Thus, the total number of molecules that are sequenced for a sample is simply a nuisance parameter.

All HTS data must be standardized and normalized after sequencing to make the different samples comparable because the total count per sample is non-informative. All of these normalizations are ratios but there are two main main approaches to identifying the denominator.  The first class of normalization uses a denominator that is internal to each sample. This can be as simple as converting the counts to proportions; i.e., by dividing the count for each feature in a sample by the total count of the sample. Derivatives of this method include the TPM (transcripts per million), FPKM( fragments per million), and RPKM (reads per kilobase per million) [@Mortazavi:2008]. In these approaches the proportion is multiplied by several constants that depend on the instrument, gene characteristics in the sample and the sequencing depth. These normalizations track closely the counts of genes determined using fluorescent hybridization methods, with typical correlations above 0.8 [@Moffitt:2016aa,@Taniguchi:2010aa].  The other internal-only normalization is the CLR (centred log-ratio) transform[@aitchison1982]. Here, the count of each feature in each sample is divided by the geometric mean of the counts in the sample with a logarithm being taken of the resulting ratio. In this case, a small pseudo-count or prior is first applied since a log(0) is undefined. The second class of normalization uses a denominator that is determined from data external to the sample being normalized, but that is still derived only from the sequencing data itself. Normalizations such as the RLE (relative log expression)[@Anders:2010], TMM (trimmed mean of M values)[@Robinson:2010a] and CSS (cumulative sum scaling)[@Paulson:2013aa] methods assume that a majority of the features are invariant and that a relative scale can be determined for each sample by dividing by a correction factor. These normalizations specifically normalize the read counts for each feature to a common total so that the counts can be compared between samples directly. In particular the TMM normalization was intended to account for asymmetry in occurrance or abundance in one group relative to the other, but assumes that the majority of features are invariant [@Robinson:2010a]. Note that the CLR normalization also assumes that a majority of the parts are invariant [@Wu2021].

A final complication arises during analysis when logarithmic transformations may or may not be applied, and the CLR is the only transformation that explicitly uses log-transformed data as the starting point. Several groups have pointed out that apparent differences between the behaviour of individual normalizations [@Paulson:2013aa, @Skinnider:2019vc] can be largely explained by differential application of logarithms [@Costea:2014aa, @erb:pc2020]. Indeed in many datasets a logarithm of one of these transforms exhibits behaviours that is similar to the logarithm of one or more other transforms [@GloorAJS2023]. Finally, each normalization is a point estimate of the appropriate denominator and so an additional assumption is that the denominator is a good estimator of the measure.

These tranformations along with the process of sequencing itself remove all but the relative variation between samples in the dataset. Nevertheless, data that are generated by sequencing come from systems where scale is usually important and may be a confounding variable [@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than do non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria often have different growth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance [@Zozaya:2010], and the composition is dramatically different as well [@Ravel:2010,@Hummelen:2010]. Thus, the full description of these systems includes both relative change (composition) and absolute abundance (scale) but we can access only the composition.

It is enlightening to examine the dispersion or variance of the data as counts and as the logarithm of counts. The counts of data derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010] as seen in Panel A of Figure 1 and this is why most tools model the counts with this distribution. However, the actual analysis is usually performed on the logarithm of the counts [Robinson:2010, Love:2014aa], or on the logarithm of the clr values[@fernandes:2013] and the mean-dispersion distribution of the logarithm of the counts is quite different as shown in panel B of Figure 1 and as noted elsewhere [@fernandes:2013, @Love:2014aa].

```{r dispersion, echo=F,warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as both counts and as logarithms of counts. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion, here the standard error (SE) of the mean, is very different when the data are log-transformed first. In addition, the amount of dispersion reaches a minimum at moderate count values. The red line in each panel shows the loess line of fit to the mid-point of the distributions. The dashed orange line in panel A is the line of equivalence, and in panel B is the minimum y value." }
library(DESeq2)
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
#library(ALDEx2)

@@

par(mfrow=c(1,2))
plot(log2(yst.mn), log2(yst.v), pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean count', ylab='Variance')
def <- data.frame(log2(yst.mn),log2(yst.v))
lines(lowess(def, f=0.1), col=2, lwd=2)
title('A: Mean vs. Variance', adj=0, line= 0.8)
abline(0,1, col=rgb(1,0.6,0,0.5), lwd=2, lty=2)

l.df <- data.frame(yst.mn.log, res@listData$lfcSE)
plot(l.df, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean log2(count)', ylab='SE log2(count)',ylim=c(0,0.4))
lines(lowess(l.df), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
title('B: log2 Mean vs. log2 Variance', adj=0, line= 0.8)

By not accounting for scale we are usually identifying statistically significant changes that are biologically meaningless but are found because of the artificially low dispersion.  As one example, recent guidance suggests \emph{using only a  small number of samples} for bulk RNA-sequencing because  only a few samples is needed to identify the majority of statistically significant features found with a large sample size [@Schurch:2016aa]. This popularized the useage of both p-values and log2 fold change between groups as dual cutoffs when determining significance. However, it is also possible to fail to identify true positive differences because of scale [@nixon2023scale].

Clearly we need some way to account for the variation in, and effect of, the scale of the underlying systems [@nixon2023scale] when determining differential abundance (DA) using high throughput sequencing. While normalizaitons attempt to correct for differences in relative counts, no tool accounts for differences in the intrinsic or extrinsic scale of the underlying system. Moreover, while we cannot measure scale directly, we can estimate its effect post-hoc on results.

# Results - examples:

We can model the conundrum of scale by assuming that the true values from the system that we want to measure are counts  composed of both compositional and total values, and that their product is the full scaled system that fits the equation: \(W = W^{\parallel}W^{\perp}\). If the true data have $D$ features (genes, taxa, etc) and $N$ samples, then  $W$ is a $D \times N$ matrix of values and we can denote the true  count of a feature in a sample as \(W_{dn}\). However, sequencing only does not allow us to access \(W^{\parallel}\) directly, but instead returns only a single estimate of the composition for each feature \(Y^{\parallel}_{dn}\). This single value is not definitive since different technical replicates  return different numbers for \(Y^{\parallel}_{dn}\) because of sampling variation [@Marioni:2008,@fernandes:2013].

We can use Bayesian techniques to estimate both the sampling variation of \(W^{\parallel}\) and the scale uncertainty of \(W^{\perp}\) by Monte Carlo sampling with the ALDEx2 R package. ALDEx2 generates pseudo technical replicates of \(Y^{\parallel}\) by Monte Carlo sampling from a Dirichlet distribution to produce a posterior distribution for \(Y^{\parallel}\) that is  an estimate of \(W^{\parallel}\)[@fernandes:2013,@gloorAJS:2016]. Then a CLR transformation is applied to each Monte Carlo sample to produce a distribution of log-ratio transformed values. Crucially, the denominator used for the CLR can be seen as a point estimate of the scale of the system [@nixon2023scale]. Thus, modifying ALDEx2 by replacing the point estimate for the CLR denominator with Monte Carlo sampled values from a log-Normal distribution gives a distribution of credible denominators. Thus, the CLR values now are credible estimates for both  \(W^{\parallel}\) and \(W^{\perp}\). This allows the scaled CLR distribution to be used to estimate the effect of scale on the output and so make \(Y \sim W\) under the assumption of the scale uncertainty. Below we use two example datasets to show how including scale uncertainty can be used to make inferences about differentially abundant features more reliable.

The first dataset is a highly replicated yeast transcriptome where one condition is wild-type and the other has a snf-1 gene knocK0ut[@Gierlinski:2015aa]. Yeast deficient for snf-1 grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa]. This dataset has been used to argue that only a small number of replicates need to be used to identity differentially abundant genes and that different tools should be used for datasets with different sample sizes because the tools have different intrinsic statistical power [@Schurch:2016aa]. This guidance runs counter to standard statistical practice where power is intrinsically linked to sample size [@Halsey:2015aa], yet is entrenched in all fields that use HTS as an experimental readout.

Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a FDR of 0.01 ;`r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts. Applying the rule of thumb of at least a $2^{1.4}$ fold change reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.t$we.eBH < 0.01 & abs(x.e$diff.btw) > 1.4))` for ALDEx2. Clearly, the necessary assumption of most features being invariant is not justified.

The very low dispersion estimate for most of the features  arises because  scale variation in the underlying data has been removed through sequencing and normalization. The actual scale of the data is unaccessible post-sequencing but we can estimate the effect of scale on the output. To do this, we add uncertainty to the denominator that is used to calculate the log-ratio of the samples, and combine this with the posterior probabilities that ALDEx2 calculates on a per-feature basis [@fernandes:2013]. Scale uncertainty is incorporated using the `gamma` parameter that controls the amount uncertainty being included when we call either `aldex()`, or `aldex.clr()`. The `ALDEx2` package contains a sensitivity analysis function, `aldex.senAnalysis()`, that can be used to explore the effect of different amounts of scale uncertainty. In practice we suggest that a `gamma` parameter between 0.5 and 1 is realistic for most experimental designs.

```{r plot1, echo=F, fig.cap="Effect and volcano plots for unscaled and scaled transcriptome analysis. DESeq2 or ALDEx2 were used to conduct a differential abundance (DA) analysis on the yeast transcriptome dataset. The results were plotted to show the relationship between difference and dispersion (effect plot) or difference and the Benjamini-Hochberg corrected p-values (volcano plot). Panels A,B,D,E are for the unscaled analysis, and Panels C,F are for the scaled analysis. Each point represents the values for one transcript, with the color indicating if that transcript was significant in the scaled analysis and unscaled analysis (red) or in the unscaled analysis only (orange). Points in grey are not statistically signficantly different with any analysis. The horizontal dashed lines represent a log2(difference) of 1.4, which is a commonly applied cutoff when the majority of features are statistically significant."}
par(mfrow=c(2,3))
plot(res@listData$lfcSE*sqrt(ncol(yst)), res@listData$log2FoldChange, xlim=c(0,5), 
  col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference')
title('A: DESeq2 effect', adj=0, line= 0.8)
points(res@listData$lfcSE[sig.des]*sqrt(ncol(yst)),
  res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), 
  pch=19, cex=0.5)
points(res@listData$lfcSE[sig.s.ald]*sqrt(ncol(yst)),
  res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), 
  pch=19, cex=0.5)
abline(h=c(-1.4,1.4), lty=2, col='grey')

@@

The second example dataset is a vaginal metatranscriptome dataset used in [@Macklaim:2018aa,@Wu2021], where we are comparing gene expression in bacteria collected from healthy (H) and BV-affected women.  In this environment, both the relative abundance of species between groups is different as is the gene expression levels within a species [@macklaim:2013]. We further expect that the total number of bacteria is about 10X more in BV than in H [@Zozaya:2010]. Thus, this is  an extremely challenging environment to determine differential abundance. Indeed, the accepted method to analyze vaginal metatranscriptomes is to conduct a taxon by taxon analysis rather than a system-wide analysis [@macklaim:2013;@Denge00262-18;@Fettweis:2019aa] because a pooled analysis unexpectedly identifies many housekeeping genes as being differentially abundant between groups.

@@

Figure 3A shows an effect plot [@gloor:effect] of the data where reads are grouped by function, corresponding approximately to orthologous proteins regardless of the organism of origin. Each point represents one of the 3728 functions, and we can see that there are many more functions represented in the BV group (bottom) than in the healthy group (top). This is because the \textit{Lactobacilli} that dominate a healthy vaginal microbiome have reduced genome content relative to the anaerobic organisms that dominate in BV, because  there is a greater diversity of organisms in BV than in H samples and because the BV condition has at least an order of magnitude more bacteria than does the H condition.

 We can  see that there are a large number of functions that are shared between the two groups (Figure 3A), and inspection shows that these largely correspond to core metabolic functions that would not be expected  to contribute to differences in ecosystem behaviour. As a proxy for housekeeping functions the core ribosome functions (blue) shows that their mean location is not centred on 0.The major group of these housekeeping functions is located off the line of no difference (being approximately located at +1.5) and not surprisingly have among the lowest dispersion in the dataset. Nevertheless, they are identified as differentially abundant (red) along with many others. While changes in the abundance of housekeeping functions is a useful proxy for relative abundance of species in the environment, they tell us nothing about the functional capacity of the two groups because these are functions in common to every organism. Of more interest is determining the functions that are different between groups that are unique or over-expressed in one group relative to the other.

@@

```{r makeScaleMatrix, echo=T}
aldex.makeScaleMatrix <- function(gamma, diff, conditions){
  ## new scale model
  # mu1 is the base relative variance
  # mu2 is the other relative variance
  # gamma is the dispersion paramenter
  gamma=gamma
  mu1 = 1 # set base to 1
  mu2 = 1 + diff # set other to adjust this until the housekeeping is centred

@@

Up to this point, scale uncertainty has been applied uniformly to both conditions, but the scale adjustment can be applied to each condition, or even each sample independently through a custom scale matrix. This can be done quite simply with the `aldex.makeScaleMatrix()` function which produces a matrix of scale uncertainties that are distinct for each group. Applying a differential scale of 0.15, or 15% of the base scale now moves the housekeeping functions to the midline of no difference. Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant because ... [HELP]. Note that this identifies a significant number of functions that are differentially up in BV that were formerly classed as not different without scale, or when only a uniform  scale was applied. These former false negatives are noted in orange in each panel. Inspection of the functions shows that these are largely missing from the Lactobacillus species and so should actually be captured as differentially abundant. Thus, applying differential scale allows us to distinguish between both false positives (housekeeping functions in cyan) and false negatives (orange functions) even in a very difficult to analyze dataset.

Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant.

>>>>

---
title: "Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
"
shorttitle: "Scale ALDEx2"
author:
- name: Greg Gloor, Michelle Pistner Nixon, Justin Silverman
  affiliation: Dep't of Biochemistry, University of Western Ontario, Penn State
  email: ggloor@uwo.ca
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  BiocStyle::pdf_document: default
package: ALDEx2
abstract: |
    Introduction to scale simulation and FDR correction with ALDEx2.
---

@@

High throughput sequencing (HTS) is a universal tool to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a fixed-number sample from the environment, making a library, multiplexing (merging) multiple libraries and applying a fixed number of molecules to the flow cell. In essence it is a poll of the environment that is mixed with other polls and then a poll of the mixture is taken. It should be clear that the total number of molecules sequenced is driven by the capacity of the instrument and not by the number of molecules in the sampled environment.

Data that are generated by sequencing come from systems where scale is usually important and may be a confounding variable [@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than do non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria often have different growth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance [@Zozaya:2010], and the composition is dramatically different as well [@Ravel:2010;@Hummelen:2010]. Thus, the full description of these systems includes both relative change (composition) and absolute abundance (scale) but we can access only the composition directly.

More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix, then the elements of the matrix can be decomposed into its composition (relative information) and its scale (total sum). We can formally state this as \(W = W^{\parallel} \times W^{\perp}\). If the matrix has \(N\) samples and \(D\) parts, then we can uniquely identify \(W_{dn}\) as the \(d\)^{th} feature in sample \(n\).  Any data that we collect by sequencing is thought to contains only proportional information and we can represent the corresponding sequenced matrix as \(Y\). Compositional data analysis makes the strong assumption  that \(Y_{n} = Y^{\parallel}_{n} = W^{\parallel}_{n}\); in other words that relative information is the only information available post-sequencing. However, the values after sequencing are only estimates of the true underlying values that we want to estimate and different technical and biological replicates vary. Note that under this strong assumption no corresponding estimate is obtained for the scale of the sample \(Y^{\perp}_{n}\).

One issue that was realized very early was that if the goal was to compare one sample to another then the output from HTS needed to be normalized in order to make the samples commensurate and to correct any minor asymmetries in the data from different samples. A large number of normalizations were developed that depended on the data source. These include rarefaction[@Hughes:2005tu], proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million[@Mortazavi:2008; @wagner:tpm]), the relative log expression (RLE)[@Anders:2010] and trimmed mean of M values (TMM)[@Robinson:2010a], and the centered log ratio (CLR)[@aitchison1982]. With the exception of rarefaction, all of these normalizations are ratios with the major differences between approaches being how the denominator is chosen and whether the ratio is always log transformed or not.

Recently, [,@nixon2023scale] showed that the denominator used to normalize \(Y^{\parallel}_{n}\) was an estimate of the scale of the system. In the context of compositional data analysis, this group proved that the geometric mean of \(Y^{\parallel}_{n}\) was a biased estimate of \(W^{\perp}_n\). This realization opened the possibility that \(Y^{\perp}_{n}\) could be made a less biased estimate by adding uncertainty when the geometric mean of \(Y^{\parallel}_{n}\)  was calculated. In terms of information theory, the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Information \(I\) (supplemental figure XX), suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional. This makes intuitive sense because  underlying systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim I_n\).

The realization that a strong assumption about scale is built into each denominator explains why HTS data are notoriously fickle as analysis of the same dataset with different workflows, tools and assumptions giving widely different results when benchmarked[@Thorsen:2016aa;@Quinn:2018aa;@Gardner:2019aa]. This can be understood as the result of unacknowledged bias about the scale estimate which inevitably leads to large Type 1 error rates as sample sizes increase [@Gustafson2015]. Thus, the realization that the denominator is a scale estimate opens up the possibility that we can supply alternate scale estimates and use this to determine which features are likely to be robust to differential abundance analysis even when taking scale into account.

The ALDEx2 R package [@fernandes:2013] represents a general purpose toolbox for Bayesian estimation of HTS datasets. ALDEx2 was designed originally to convert point estimates of \(Y^{\parallel}_{dn}\) into posterior distributions through Monte-Carlo sampling from the Dirichlet distribution, where the posterior more accurately represented variation on a per-sample basis[@gloorAJS:2016]. Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally  expected values and confidence intervals are reported across the Monte-Carlo replicates. The CLR calculation uses as the denominator the geometric mean of each Monte-Carlo replicate of \(Y^{\parallel}_{dn}\) with the geometric mean representing a point estimate of scale, i.e. of \(Y^{\perp}_{n}\). We  now include a posterior distribution of the scale into this process by sampling \(Y^{\perp}_n\) from a log-Normal distribution and using these as the denominator. The scale estimates can be calculated on a per-experiment, per-group or even a per-sample basis.

An advantage of incorporating scale is that the analysis can be made much more robust and that differences in scale can be accounted for explicitly. The examples below show how incorporating scale provides robust and interpretable differential abundance estimates in several different datasets.

# Results

The first dataset is a highly replicated yeast transcriptome where one condition is wild-type and the other has a snf-1 gene knockout[@Gierlinski:2015aa]. Yeast deficient for snf-1 grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa]. This dataset has been used to argue that only a small number of replicates need to be used to identity differentially abundant genes and that different tools should be used for datasets with different sample sizes because the tools have different intrinsic statistical power and Type 1 and Type 2 errors [@Schurch:2016aa]. This guidance runs counter to standard statistical practice where power is intrinsically linked to sample size [@Halsey:2015aa], yet the concept of sample-size independent power is entrenched in all fields that use HTS as an experimental readout. Increasing false precision with increasing sample size can be understood as the result of unacknowledged bias brought about because of false certainty as to the scale of the data. Thus, by adding scale uncertainty we can be more confident that the analysis is not converging on a precise but biased estimate of scale [@Gustafson2015].

We start by examining the dispersion or variance of the data as counts and as the logarithm of the normalized counts as calculated by DESeq2 (log(RLE)) and by ALDEx2 (clr). The actual counts of data derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010] as seen in Panel A of Figure 1. This relationship is why most tools model the counts with the Negative Binomial distribution[@polyester:2016], and use this distribution as the basis for batch correction[@Zhang:2020ab]. However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [Robinson:2010, Love:2014aa], or on  the clr values[@fernandes:2013] both of which are  log-ratios. The mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013, @Love:2014aa].

```{r dispersion, echo=F,warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,  as logarithms of counts, and as clr values. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first. Panel C shows the equivalent values calculated by ALDEx2. The expected clr value for each transcipt are plotted vs. the expected dispersion. In panels B and C the amount of dispersion reaches a minimum at moderate  values. The red line in each panel shows the loess line of fit to the mid-point of the distributions. The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value. The dashed grey line in panels B and C represent the maximum of the lowest 10% dispersion." }
library(DESeq2)
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
#library(ALDEx2)

@@

par(mfrow=c(1,3))
plot(log2(yst.mn), log2(yst.v), pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean count', ylab='Variance')
#points(log2(yst.mn[sig.des]), log2(yst.v[sig.des]), pch=19, cex=0.5, col='orange')
def <- data.frame(log2(yst.mn),log2(yst.v))
lines(lowess(def, f=0.1), col=2, lwd=2)
title('A: Mean vs. Variance', adj=0, line= 0.8)
abline(0,1, col=rgb(1,0.6,0,0.5), lwd=2, lty=2)

l.df <- data.frame(yst.mn.log, res@listData$lfcSE)
plot(l.df, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean log2(count)', ylab='SE log2(count)',ylim=c(0,0.4))
lines(lowess(l.df, f=0.25), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.0275, col='grey', lty=2)
title('B: log2 Mean vs. log2 Variance', adj=0, line= 0.8)

l.df <- data.frame(x.e$rab.all, x.e$diff.win)
plot(x.e$rab.all, x.e$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='E(clr)', ylab='E(dispersion)')
#points(x.e$rab.all[sig.ald], x.e$diff.win[sig.ald], col='orange', cex=0.4)
lines(lowess(l.df, f=.1), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.208, col='grey', lty=2)
title('C: E(clr) vs. E(dispersion)', adj=0, line= 0.8)
```

Here we can see that while the variance or dispersion always increases with increasing raw read count, it actually decreases when measured on the log-ratio transformed data[@fernandes:2013; @Love:2014aa] and reaches a minimum at some mid-point of the distribution. This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes. This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation[@Taniguchi:2010aa]. Furthermore, the actual amount of dispersion is very small and for many transcripts is almost negligible. To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives. The work of [,@nixon2023scale] suggests that this is because the biased estimate of \(Y^{\perp}\) that is  calculated leads to a false estimate of the scale. Thus, the actual variation of highly expressed genes is not captured accurately by current approaches to analysis and supports the idea that HTS have unacknowledged bias because scale is not taken into account.

Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a FDR of 0.01 ;`r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts. Clearly, the necessary assumption of most features being invariant is not justified. Applying the rule of thumb of at least a $2^{1.4}$ fold change [@Schurch:2016aa] reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.t$we.eBH < 0.01 & abs(x.e$diff.btw) > 1.4))` for ALDEx2.

```{r plot1, echo=F, fig.cap="Effect and volcano plots for unscaled and scaled transcriptome analysis. DESeq2 or ALDEx2 were used to conduct a differential abundance (DA) analysis on the yeast transcriptome dataset. The results were plotted to show the relationship between difference and dispersion (effect plot [@gloor:effect]) or difference and the Benjamini-Hochberg corrected p-values (volcano plot[@Cui:2003aa]). Panels A,B,D,E are for the unscaled analysis, and Panels C,F are for the scaled analysis. Each point represents the values for one transcript, with the color indicating if that transcript was significant in the scaled analysis and unscaled analysis (red) or in the unscaled analysis only (orange). Points in grey are not statistically signficantly different with any analysis. The horizontal dashed lines represent a log2(difference) of 1.4, which is a commonly applied cutoff when the majority of features are statistically significant."}
par(mfrow=c(2,3))
plot(res@listData$lfcSE*sqrt(ncol(yst)), res@listData$log2FoldChange, xlim=c(0,5), 
  col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference')
title('A: DESeq2 effect', adj=0, line= 0.8)
points(res@listData$lfcSE[sig.des]*sqrt(ncol(yst)),
  res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), 
  pch=19, cex=0.5)
points(res@listData$lfcSE[sig.s.ald]*sqrt(ncol(yst)),
  res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), 
  pch=19, cex=0.5)
abline(h=c(-1.4,1.4), lty=2, col='grey')

@@

The very low dispersion estimate for most of the features  arises because  scale variation in the underlying data has been removed through sequencing and normalization. While the actual scale of the data is inaccessible post-sequencing  we can estimate the effect of scale on the output. To do this, we add uncertainty to the denominator that is used to calculate the log-ratio of the samples [@nixon2023scale], and combine this with the posterior probabilities that ALDEx2 calculates on a per-feature basis [@fernandes:2013]. Scale uncertainty is incorporated using the `gamma` parameter that controls the amount uncertainty being included when we call either `aldex()`, or `aldex.clr()`. The `ALDEx2` package contains a sensitivity analysis function, `aldex.senAnalysis()`, that can be used to explore the effect of different amounts of scale uncertainty. In practice we suggest that a `gamma` parameter between 0.5 and 1 is realistic for most experimental designs.

The second example dataset is a vaginal metatranscriptome dataset used in [@Macklaim:2018aa,@Wu2021], where we are comparing gene expression in bacteria collected from healthy (H) and BV-affected women.  In this environment, both the relative abundance of species between groups is different as is the gene expression levels within a species [@macklaim:2013]. We further know that the total number of bacteria is about 10X more in BV than in H [@Zozaya:2010]. Thus, this is  an extremely challenging environment to determine differential abundance. Indeed, the accepted method to analyze vaginal metatranscriptomes is to conduct a taxon by taxon analysis rather than a system-wide analysis [@macklaim:2013;@Denge00262-18;@Fettweis:2019aa] because a pooled analysis unexpectedly identifies many housekeeping genes as being differentially abundant between groups.

@@

Figure 3A shows an effect plot [@gloor:effect] of the data where reads are grouped by function, corresponding approximately to grouping orthologous sequences regardless of the organism of origin. Each point represents one of the 3728 functions, and we can see that there are many more functions represented in the BV group (bottom) than in the healthy group (top). This is because the \textit{Lactobacilli} that dominate a healthy vaginal microbiome have reduced genome content relative to the anaerobic organisms that dominate in BV, because  there is a greater diversity of organisms in BV than in H samples and because the BV condition has at least an order of magnitude more bacteria than does the H condition.

 We can  see that there are a large number of functions that are shared between the two groups (boxed area in Figure 3A), and inspection shows that these largely correspond to core metabolic functions that would not be expected  to contribute to differences in ecosystem behaviour. As a proxy for housekeeping functions the core ribosome functions (blue) shows that their mean location is not centred on 0. The major group of these housekeeping functions is located off the line of no difference (being approximately located at +1.5) and not surprisingly have among the lowest dispersion in the dataset. Nevertheless, they are identified as differentially abundant (red) along with many others. While changes in the abundance of housekeeping functions is a useful proxy for relative abundance of species in the environment, they tell us nothing about the functional capacity of the two groups because these are functions in common to every organism. Of more interest is determining the functions that are different between groups because these are unique or over-expressed in one group relative to the other.

@@

```{r makeScaleMatrix, echo=F}
aldex.makeScaleMatrix <- function(gamma, diff, conditions){
  ## new scale model
  # mu1 is the base relative variance
  # mu2 is the other relative variance
  # gamma is the dispersion paramenter
  gamma=gamma
  mu1 = 1 # set base to 1
  mu2 = 1 + diff # set other to adjust this until the housekeeping is centred

@@

Up to this point, scale uncertainty has been applied uniformly to both conditions, but the scale adjustment can be applied to each condition, or even each sample independently through a custom scale matrix. This can be done quite simply with the `aldex.makeScaleMatrix()` function which produces a matrix of scale uncertainties that are distinct for each group. Applying a differential scale of 0.15, or 15% of the base scale now moves the housekeeping functions to the midline of no difference (Figure 3C). Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant because the change in scael moves the mass of the data towards the group with the highest entropy (supplemental). Note that this identifies a significant number of functions that are differentially up in BV that were formerly classed as not different without scale, or when only a uniform  scale was applied. These former false negatives are noted in orange in each panel. Inspection of the functions shows that these are largely missing from the Lactobacillus species and so should actually be captured as differentially abundant. Thus, applying differential scale allows us to distinguish between both false positives (housekeeping functions in cyan) and false negatives (orange functions) even in a very difficult to analyze dataset.
